---
title: "Principal component Analysis/Factor Analysis"
format:
  html:
    self-contained: true
editor: visual
---

-   **\#### PCA**:

-   Reduces the number of variables into a smaller set of components that still explain most of the variation in the data.

-   We have **8 workload outcomes** (mental, physical, emotional, temporal, etc.).These outcomes are **correlated** — when one goes up, others often do too.

-   Running 8 separate models is messy and makes it hard to summarize.

-   PCA (or factor analysis) lets us **combine these 8 measures into a smaller number of workload “dimensions”**.

    -   PC1 could represent **“Overall workload”** (captures mental + physical + temporal effort).

    -   PC2 could represent **“Emotional strain”** (captures stress + frustration).

-   So instead of predicting 8 scores, we predict 2–3 interpretable scores.

-   We then used patient characteristics to predict these workload scores, and designed an assignment system so that each physician receives a fair share of overall workload and emotional strain

```{r setup,echo=FALSE,message=FALSE,error=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = FALSE,      # hide R code
  message = FALSE,   # suppress messages
  warning = FALSE,   # suppress warnings (optional)
  error = FALSE      # don't show errors in output
)
```

```{r,echo=FALSE,message=FALSE,error=FALSE}
# =============================================
# Workload Factor Analysis & Score Computation
# =============================================

# libraries
library(psych)       # For factor analysis
library(GPArotation) # For rotation options
library(janitor)
library(readxl)
library(tidyverse)
```

```{r,echo=FALSE,message=FALSE,error=FALSE}
# -----------------------------
# 0. import and clean data 
# -----------------------------
# load data
PCA_data <- read_tsv("/projects/cshcd/kern26/s312369.hosp.wl/adnan/Workload/data/ModelingData.txt")

# clean column names
PCA_data <- janitor::clean_names(PCA_data)

dependent_vars <- c(
  "physical", "mental", "temporal",
  "performance", "effort", "frustration"
  # ,"time", "chart"
)

dep_data <- PCA_data %>% select(all_of(dependent_vars))
```

#### Correlation matrix

```{r}
# -----------------------------
# 1. Check correlation matrix
# -----------------------------

cor_matrix <- cor(dep_data, use = "pairwise.complete.obs")
print(cor_matrix)
```

#### Parallel analysis to decide number of factors

```{r}
# -----------------------------
# 2. Parallel analysis to decide number of factors
# -----------------------------
pa <- fa.parallel(dep_data, fa = "fa", n.iter = 100, plot = TRUE, fm = "ml", cor = "cor")  
```

#### Factor analysis with `r pa$nfact` factors

```{r}
# -----------------------------
# 3. Standardize the data
# -----------------------------
# Ensures variables on different scales are comparable
dep_data_scaled <- scale(dep_data)

# -----------------------------
# 4. Run factor analysis
# -----------------------------
# Choose number of factors suggested by parallel analysis (here nfactors=4)
fa_result <- fa(dep_data_scaled, nfactors = 4, rotate = "varimax", fm = "ml")

# -----------------------------
# 5. Inspect factor loadings
# -----------------------------
print(fa_result) #, cut = 0.3)  # hides small loadings < 0.3
# -----------------------------
# 4a. Sort factors by variance explained
# -----------------------------
fa.sort(fa_result)  # sorts factors so the first explains most variance

# -----------------------------
# 5. Inspect factor loadings
# -----------------------------
print(fa_result_sorted)      # nicely sorted
fa_result_sorted$loadings   # full sorted loadings
fa_result$loadings          # full loadings
```

#### Factor scores for each patient

```{r}
# -----------------------------
# 6. Compute factor scores for each patient
# -----------------------------
factor_scores <- factor.scores(dep_data_scaled, fa_result)$scores
factor_scores <- as.data.frame(factor_scores)
head(factor_scores)
```

```{r}
# -----------------------------
# 7. Compute a single workload score
# -----------------------------
# Option B: weighted by variance explained (SS loadings / sum)
weights <- fa_result$Vaccounted["SS loadings", ] / sum(fa_result$Vaccounted["SS loadings", ])
workload_score_weighted <- as.matrix(factor_scores) %*% weights

# Inspect final workload scores
head(workload_score_weighted)
```

-   **composite (final) workload score** is centered around 0, min workload score = `r min(workload_score_weighted)` max workload score = `r max(workload_score_weighted)`

### Principal component analysis

```{r}
# -----------------------------
# 0. import and clean data 
# -----------------------------
# load data
PCA_data <- read_tsv("/projects/cshcd/kern26/s312369.hosp.wl/adnan/Workload/data/ModelingData.txt")

# clean column names
PCA_data <- janitor::clean_names(PCA_data)

dependent_vars <- c(
  "physical", "mental", "temporal",
  "performance", "effort", "frustration"
  # ,  "time", "chart"
)

dep_data <- PCA_data %>% select(all_of(dependent_vars))
```

#### **Parallel analysis to decide number of factors**

```{r}
# -----------------------------
# 2. Parallel analysis to decide number of components
# -----------------------------
pa <- fa.parallel(dep_data, fa = "pc", n.iter = 100, plot = TRUE, fm = "ml", cor = "cor")  
```

#### PCA with `r pa$ncomp` factors

```{r}
# -----------------------------
# 3. Standardize the data
# -----------------------------
dep_data_scaled <- scale(dep_data)

# -----------------------------
# 4. Run PCA
# -----------------------------
# Suppose parallel analysis suggests 3 components
pca_result <- principal(dep_data_scaled, nfactors = 3, rotate = "varimax")

# -----------------------------
# 5. Inspect PCA loadings
# -----------------------------
print(pca_result, cut = 0.3)   # suppress loadings <0.3
pca_result$loadings            # full loadings
```

#### PCA scores for each patient

```{r}
# -----------------------------
# 6. Extract PCA scores
# -----------------------------
pca_scores <- as.data.frame(pca_result$scores)
head(pca_scores)
```

```{r}
# -----------------------------
# 7. Compute overall workload score
# -----------------------------
# Option B: weighted by variance explained
weights <- pca_result$Vaccounted["Proportion Var", ]
workload_score_weighted <- as.matrix(pca_scores) %*% weights
# Inspect final workload scores
head(workload_score_weighted)
```
